\documentclass[12pt]{article}
\usepackage{a4, fullpage}
\usepackage{bibtopic}
\usepackage{float}
\usepackage{amssymb,amsmath}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{alltt}

\begin{document}

\title{Equivalences Tutor\\BEng Final Report}

\author{Sahil Deepak Jain\\Department of Computing\\Imperial College London\\Supervisor: Dr. Fariba Sadri}
\date{\today}
\maketitle
\newpage

% Abstract

\begin{abstract}

\end{abstract}
\newpage

% Table Of Contents
\tableofcontents

\newpage
% Introduction
\section{Introduction}

\emph{Logic} refers to the study of different modes of reasoning conducted or 
assessed according to strict principles of validity. Due to Logic being one of 
the most fundamental aspects of Computer Science, it is taught to every student
pursuing a Computing degree at Imperial College London. During the first term 
in university, every Computing student is taught the Logic course, which aims to
provide the students with knowledge of the syntax and semantics of Propositional 
and Predicate logic. Students can apply this knowledge to complete equivalences
and natural deduction proofs.

A logical system is made up of three things:

\begin{enumerate}
  \item Syntax - this is the formal language specified to express different
         concepts.
  \item Semantics - this is what provides meaning to the language.
  \item Proof theory - this is a way of arguing in the language. This allows us
         to identify valid statements in the language.
\end{enumerate}

In logic, two statements are logically equivalent if they contain the same
logical content. Mendelson stated that "two statements are equivalent if they
have the same truth value in every model." This can be illustrated in the 
following example: \\ \bigskip 
Statement 1: \emph{If Sahil is a final year student, 
then he has to do an individual project} \\ \bigskip 
Statement 2: \emph{If Sahil is not 
doing an individual project, then he is not a final year student} \\ \bigskip 
As we can see, both statements have the same result in same models. When two logic 
statements are equivalent, they can be derived by each other, with the use of 
equivalences which we know to be true.

\subsection{Current Logic Tutorial Applications}

At the moment, there are a two programs which specifically built to support the
Logic course at Imperial College London. These are:

\begin{enumerate}
  \item Pandora - learning support tool designed to guide the construction of
         natural deduction proofs. 
  \item LOST - application which helps in the learning of logic semantics.
\end{enumerate}

There is no current tutorial tool for logic equivalences.

\subsection{Requirements}

The aim of this project is to provide a web based equivalence tutorial tool
with the following features:

\begin{itemize}

	\item Input start and end states of a propositional or predicate equivalence, 
         and perform the equivalence.
	\item Choose a difficulty of equivalence and practice an auto-generated
			   level.
	\item Choose between a logic tree view and a logic expression view while
				 completing a level.
	\item Give the student the ability to view previously completed levels.
	\item Give the student the ability to save uncompleted levels, which they
				 can complete at a later time.
	\item Automatically suggest a higher level to the student if they are
				 progressing well.
\end{itemize}

\newpage


% Background
\section{Background}
\subsection{Propositional Equivalences}

The first thing I had to look at were the logic equivalences which the students
are taught in their respective logic courses. These equivalences are described
in the next few subsections.

\subsubsection{And Equivalences}
\begin{enumerate}

  \item A $\land$ B $\equiv$ B $\land$ A \emph{(Commutativity of And)}

\begin{center}
  \begin{tabular}{| c | c | c | c |}
    \hline
    A & B & A $\land$ B & B $\land$ A \\ \hline
    0 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 \\
    1 & 0 & 0 & 0 \\
    1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item A $\land$ A $\equiv$ A \emph{(Idempotence of And)}

\begin{center}
  \begin{tabular}{| c | c |}
    \hline
    A & A $\land$ A \\ \hline
    0 & 0 \\
    0 & 0 \\
    1 & 1 \\
    1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item (A $\land$ B) $\land$ C $\equiv$ A $\land$ (B $\land$ C) \emph{(Associaticity of And)}

\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    A & B & C & (A $\land$ B) $\land$ C & A $\land$ (B $\land$ C) \\ \hline
    0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 0 & 0 \\
    0 & 1 & 0 & 0 & 0 \\
    0 & 1 & 1 & 0 & 0 \\
    1 & 0 & 0 & 0 & 0 \\
    1 & 0 & 1 & 0 & 0 \\
    1 & 1 & 0 & 0 & 0 \\
    1 & 1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}
\end{enumerate}

\subsubsection{Or Equivalences}
\begin{enumerate}

  \item A $\lor$ B $\equiv$ B $\lor$ A \emph{(Commutativity of Or)}

\begin{center}
  \begin{tabular}{| c | c | c | c |}
    \hline
    A & B & A $\lor$ B & B $\lor$ A\\ \hline
    0 & 0 & 0 & 0 \\
    0 & 1 & 1 & 1 \\
    1 & 0 & 1 & 1 \\
    1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}

\item A $\lor$ A $\equiv$ A \emph{(Idempotence of Or)}

\begin{center}
  \begin{tabular}{| c | c |}
    \hline
    A & A $\lor$ A \\ \hline
    0 & 0 \\
    0 & 0 \\
    1 & 1 \\
    1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item (A $\lor$ B) $\lor$ C $\equiv$ A $\lor$ (B $\lor$ C) \emph{(Associaticity of Or)}

\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    A & B & C & (A $\lor$ B) $\lor$ C & A $\lor$ (B $\lor$ C) \\ \hline
    0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 1 & 1 \\
    0 & 1 & 0 & 1 & 1 \\
    0 & 1 & 1 & 1 & 1 \\
    1 & 0 & 0 & 1 & 1 \\
    1 & 0 & 1 & 1 & 1 \\
    1 & 1 & 0 & 1 & 1 \\
    1 & 1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}
\end{enumerate}

\subsubsection{Not Equivalences}
\begin{enumerate}

  \item A $\equiv$ $\neg$$\neg$A

\begin{center}
  \begin{tabular}{| c | c | c |}
    \hline
    A & $\neg$A & $\neg$$\neg$A \\ \hline
    0 & 1 & 0 \\
    1 & 0 & 1 \\ \hline
  \end{tabular}
\end{center}

\end{enumerate}

\subsubsection{Implies Equivalences}
\begin{enumerate}

  \item A $\to$ B $\equiv$ $\neg$A $\lor$ B

\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    A & B & $\neg$A & A $\to$ B & $\neg$A $\lor$ B \\ \hline
    0 & 0 & 1 & 1 & 1 \\
    0 & 1 & 1 & 1 & 1 \\ 
    1 & 0 & 0 & 0 & 0 \\
    1 & 1 & 0 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item A $\to$ B $\equiv$ $\neg$(A $\land$ $\neg$B)

\begin{center}
  \begin{tabular}{| c | c | c | c | c | c |}
    \hline
    A & B & $\neg$B & A $\land$ $\neg$B & A $\to$ B & $\neg$(A $\land$ $\neg$B) \\ \hline
    0 & 0 & 1 & 0 & 1 & 1 \\
    0 & 1 & 0 & 0 & 1 & 1 \\
    1 & 0 & 1 & 1 & 0 & 0 \\
    1 & 1 & 0 & 0 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item $\neg$(A $\to$ B) $\equiv$ A $\land$ $\neg$B

\begin{center}
  \begin{tabular}{| c | c | c | c | c | c |}
    \hline
    A & B & $\neg$B & A $\to$ B & $\neg$(A $\to$ B) & A $\land$ $\neg$B \\ \hline
    0 & 0 & 1 & 1 & 0 & 0 \\
    0 & 1 & 0 & 1 & 0 & 0 \\
    1 & 0 & 1 & 0 & 1 & 1 \\
    1 & 1 & 0 & 1 & 0 & 0 \\ \hline
  \end{tabular}
\end{center}
\end{enumerate}

\subsubsection{If and Only If Equivalences}
\begin{enumerate}

  \item A $\leftrightarrow$ B $\equiv$ (A $\to$ B) $\land$ (B $\to$ A)

\begin{center}
  \begin{tabular} {| c | c | c | c | c | c |}
    \hline
    A & B & A $\to$ B & B $\to$ A & A $\leftrightarrow$ B & (A $\to$ B) $\land$ (B $\to$ A) \\ \hline
    0 & 0 & 1 & 1 & 1 & 1 \\
    0 & 1 & 1 & 0 & 0 & 0 \\
    1 & 0 & 0 & 1 & 0 & 0 \\
    1 & 1 & 1 & 1 & 1 & 1 \\ \hline
  \end{tabular} 
\end{center}

  \item A $\leftrightarrow$ B $\equiv$ (A $\land$ B) $\lor$ ($\neg$A $\land$ $\neg$B)

\begin{center}
  \begin{tabular} {| c | c | c | c | c | c | c |}
    \hline
    A & B & A $\land$ B & $\neg$A & $\neg$B & A $\leftrightarrow$ B & (A $\land$ B) $\lor$ ($\neg$A $\land$ $\neg$B) \\ \hline
    0 & 0 & 0 & 1 & 1 & 1 & 1 \\
    0 & 1 & 0 & 1 & 0 & 0 & 0 \\
    1 & 0 & 0 & 0 & 1 & 0 & 0 \\
    1 & 1 & 1 & 0 & 0 & 1 & 1 \\ \hline
  \end{tabular} 
\end{center}
\end{enumerate}

\subsubsection{De Morgan Equivalences}

\begin{enumerate}

  \item $\neg$(A $\land$ B) $\equiv$ $\neg$A $\lor$ $\neg$B

\begin{center}
  \begin{tabular}{| c | c | c | c |}
    \hline
    A & B & $\neg$(A $\land$ B) & $\neg$A $\lor$ $\neg$B \\ \hline
    0 & 0 & 1 & 1 \\
    0 & 1 & 1 & 1 \\ 
    1 & 0 & 1 & 1 \\
    1 & 1 & 0 & 0 \\ \hline
  \end{tabular}
\end{center}

  \item $\neg$(A $\lor$ B) $\equiv$ $\neg$A $\land$ $\neg$B

\begin{center}
  \begin{tabular}{| c | c | c | c |}
    \hline
    A & B & $\neg$(A $\lor$ B) & $\neg$A $\land$ $\neg$B \\ \hline
    0 & 0 & 1 & 1 \\
    0 & 1 & 0 & 0 \\ 
    1 & 0 & 0 & 0 \\
    1 & 1 & 0 & 0 \\ \hline
  \end{tabular}
\end{center}
\end{enumerate}

\subsubsection{Distributibivity of And and Or}
\begin{enumerate}

  \item A $\land$ (B $\lor$ C) $\equiv$ (A $\land$ B) $\lor$ (A $\land$ C)

\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    A & B & C & A $\land$ (B $\lor$ C) & (A $\land$ B) $\lor$ (A $\land$ C) \\ \hline
    0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 0 & 0 \\
    0 & 1 & 0 & 0 & 0 \\
    0 & 1 & 1 & 0 & 0 \\
    1 & 0 & 0 & 0 & 0 \\
    1 & 0 & 1 & 1 & 1 \\
    1 & 1 & 0 & 1 & 1 \\
    1 & 1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}

  \item A $\lor$ (B $\land$ C) $\equiv$ (A $\lor$ B) $\land$ (A $\lor$ C)

\begin{center}
  \begin{tabular}{| c | c | c | c | c |}
    \hline
    A & B & C & A $\lor$ (B $\land$ C) & (A $\lor$ B) $\land$ (A $\lor$ C) \\ \hline
    0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 0 & 0 \\
    0 & 1 & 0 & 0 & 0 \\
    0 & 1 & 1 & 1 & 1 \\
    1 & 0 & 0 & 1 & 1 \\
    1 & 0 & 1 & 1 & 1 \\
    1 & 1 & 0 & 1 & 1 \\
    1 & 1 & 1 & 1 & 1 \\ \hline
  \end{tabular}
\end{center}
\end{enumerate}
\subsection{Equivalence Examples}
\begin{enumerate}

  \item $\neg$(p $\to$ q) $\equiv$ p $\land$ $\neg$q
  \begin{enumerate}
    \item $\neg$(p $\to$ q)
    \item $\neg$($\neg$p $\lor$ q) \hfill a, Implication rule
    \item $\neg$$\neg$p $\land$ $\neg$q \hfill b, De Morgan's law
    \item p $\land$ $\neg$q \hfill c, Double Negation rule
  \end{enumerate}
  
  \item $\neg$((p $\land$ q) $\to$ r) $\equiv$ (p $\land$ q) $\land$ $\neg$r
  \begin{enumerate}
    \item $\neg$((p $\land$ q) $\to$ r)
    \item $\neg$($\neg$(p $\land$ q) $\lor$ r) \hfill a, Implication rule
    \item $\neg$$\neg$(p $\land$ q) $\land$ $\neg$ r \hfill b, De Morgan's law
    \item (p $\land$ q) $\land$ $\neg$r \hfill c, Double Negation rule
  \end{enumerate}

  \item $\neg$(p $\lor$ $\neg$q) $\lor$ ($\neg$p $\land$ $\neg$q) $\equiv$ $\neg$p
  \begin{enumerate}
    \item $\neg$(p $\lor$ $\neg$q) $\lor$ ($\neg$p $\land$ $\neg$q)
    \item $\neg$(p $\lor$ $\neg$q) $\lor$ $\neg$(p $\land$ q) \hfill a, De Morgan's law
    \item $\neg$[(p $\lor$ $\neg$q) $\land$ (p $\land$ q)] \hfill b, De Morgan's law
    \item $\neg$[p $\lor$ ($\neg$q $\land$ q)] \hfill c, Distributivity
    \item $\neg$(p $\lor$ $\bot$) \hfill d, And rule
    \item $\neg$p \hfill e, Or rule
  \end{enumerate}
\end{enumerate}

\newpage

\section{Design and Implementation Overview}
\subsection{Lexer and Parser}
As we have had previous experiences with writing lexers and parsers, this was not
a hard part of the development process. With the use of ANTLR, which is a lexer
and parser generator, I have achieved a fully functional parser for propositional
logic, which can easily be extended to account for predicate logic in future
development.

\subsection{Logic Tree Structure}
To generate the logic tree, I chose not to use the ANTLR generated Abstract Syntax
Trees. Instead, I wrote a tree structure of my own. The parser grammar can be seen
in the appendix, which would help with the understanding of the tree structure as well.

\subsection{Graphical User Interface}
I am now starting to develop the GUI for the application. As the main implementation
of the application is in Java, I have chosen to use the Swing library to develop
the GUI.

\newpage

% Project Plan
\section{Project Plan}
As we are building a completely new application, we have to start from the
very basics of what to do. The inital stages of the project consisted of the
following:
\begin{enumerate}
	\item Meeting up with project supervisor to discuss what she would like as
				 an end result of the project.
	\item Reading through equivalence notes from previous years.
	\item Talk to a few first year students to know where they found problems
				 with equivalences.
	\item Research about the best way to model logic expressions so they could
   			 be easily manipulated in future development.
\end{enumerate}

Once these steps had been carried out, I had a rough idea of where to begin 
the development aspect of the application. After thorough research, I had realised
that the best way to model the logic expressions was to use a tree structure, which
was suggested by Fariba as well. To create the tree structure, I chose to use
ANTLR. This stage consisted of:

\begin{enumerate}
	\item Writing the Lexer rules.
	\item Writing the Parser rules.
	\item Research more about the Abstract Syntax Tree (AST) returned by ANTLR.
\end{enumerate}

After this further research, I realised that using the ANTLR generated ASTs would
not be the most convenient tree structure to use in the long run, so I decided
to create a tree structure of my own.

Future steps include (in order of implementation):
\begin{enumerate}
	\item Designing the GUI
	\item Developing the GUI
	\item Writing the equivalence rules which alter the tree structure
	\item Writing the logic expression generator
	\item Writing the progress tracker
  \item Extending the application to support predicate equivalences
\end{enumerate}
\newpage

%Evaluation Plan
\section{Evaluation Plan}
\subsection{User Feedback}
As this is a tutorial tool, the best way to evaluate the application is to see
if the people who use it find it useful in learning equivalences. There are a number
of aspects which would need to be evaluated. These are discussed below.

\subsubsection{Ease of Use}
For a tutorial tool, or any application, to be a good application, it has to be
user friendly. When talking about this application specifically, we have to
consider many different areas when talking about ease of use:
\begin{enumerate}
  \item Inputting start and end states for an equivalence - this would be the
         first step for a user to complete an equivalence, unless they are using
         the auto-generated equivalences. Input of logic expressions should be
         simple and fast. If this aspect of the tutorial tool is not easy to use,
         the user would not want to waste time on the application as it would be
         too time consuming.
  \item Solving equivalences - this part is the main feature of the application.
         This aspect of the application should be very simple and not require many
         clicks or lots of typing, and should provide an easy way of undoing rules
         that the user might have applied. Both the tree method and the expression
         method of solving equivalences should be very easy and fast to use.
  \item Saving and opening uncompleted equivalences - this feature of the tutorial
         tool would be essential for a user who might have a specific equivalence 
         too hard to complete at a given point, so they can return to it in the
         future when they are more comfortable with equivalences. It should
         be easy to see the start and end states of the uncompleted levels, and
         the users should be allowed to sort it by date or difficulty if need be.
  \item Generating new levels - if the user does not have any equivalences to 
         enter themselves, they can generate levels of their chosen level, or
         generate a level which the application thinks is suitable. This again,
         should be done within a few clicks of the mouse, as the user would not
         want to waste time on it.
  \item Viewing Progress Report - the user can keep track of their progress with
         equivalences. This should be shown with easy to read and understand
         statistics of the user's progress. 
\end{enumerate}

\subsubsection{Difficulty}
Every tutorial tool needs to give the user tougher challenges as they progress.
In this application, we keep track of certain statistics of each user, such as
the number of completed levels, number of incomplete levels, average steps needed
to complete a level, number of easy/medium/hard levels completed, etc. 
As there is a level generator, we need to keep track of these statistics, and
after analysing the data, the application would generate a suitable level for the
user.
To test this aspect of the application, we would have a few users who would complete
generated levels, and with time, they would see the levels getting harder and harder.
After a certain time, we would ask the users if they felt that the levels were getting
harder and if they felt under-challenged or over-challenged. This would provide the
basis of evaluation of the level generation aspect of the tutorial tool.

\subsection{Testing}
Thorough testing is essential to any application. Many tests will be written to
check that no completed part of the application is broken once new parts are added,
making sure that every feature of the application works.
\newpage

\section{Appendix}

\subsection{Lexer Grammar}

\begin{verbatim}
lexer grammar LogicLexer;

options {
  language = Java;
}

@header {
  package eqtutor;
}

WHITESPACE:			( '\t' | ' ' | '\r' | '\n' | '\u000C' )+ { $channel = HIDDEN; };

AND	:						'&';
OR	:						'|';
IFTHEN	:				'->';
IFF 	:					'<>';
NOT 	:					'!';

LPAREN  :				'(';
RPAREN  :				')';

ID	:						('A'..'Z'|'a'..'z') ('A'..'Z'|'a'..'z'|'_')*;
\end{verbatim}

\newpage

\subsection{Parser Grammar}

\begin{verbatim}
parser grammar LogicParser;

options {
  tokenVocab = LogicLexer;
}

@header {
  package eqtutor;

  import java.util.LinkedList;
  import java.util.List;
  import AST.*;
}

@members {
  private boolean hasFoundError = false;

  public void displayRecognitionError(String[] tokenNames, RecognitionException e) {
    hasFoundError = true;
  }

  public boolean hasFoundError() {
    return this.hasFoundError;
  }
}

program returns [AST tree]
  : e = iffexpr {$tree = new AST(new ASTProgramNode($e.node));} EOF
  ;

iffexpr returns [ASTPropositionalNode node]
  : ifthen = ifexpr {$node = $ifthen.node;} 
	  (IFF iff = iffexpr {$node = new ASTIffNode($ifthen.node, $iff.node);})*
  ;
  	
ifexpr returns [ASTPropositionalNode node]
  : or = orexpr {$node = $or.node;} 
		(IFTHEN ifthen = ifexpr {$node = new ASTIfThenNode($or.node, $ifthen.node);})*
  ;

orexpr returns [ASTPropositionalNode node]
  : and = andexpr {$node = $and.node;} 
		(OR or = orexpr {$node = new ASTOrNode($and.node, $or.node);})*
  ;

andexpr returns [ASTPropositionalNode node]
  : not = notexpr {$node = $not.node;} 
		(AND and = andexpr {$node = new ASTAndNode($not.node, $and.node);})*
  ;

notexpr returns [ASTPropositionalNode node]
  : NOT not = notexpr {$node = new ASTNotNode($not.node);}
  | id = identifier {$node = $id.node;}
  ;

identifier returns [ASTPropositionalNode node]
  : ID {$node = new ASTIdentifierNode($ID.text);}
  | LPAREN iffexpr RPAREN {$node = $iffexpr.node;}
  ;
\end{verbatim}

\end{document}
